Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-20 reanalysis-4-outcome_analysis.qmd
fix

# Create interaction plot
p1 <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds, 
                            color = Treatment, group = Treatment)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.1, size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "Three-Way Interaction: Lapse × DR Severity × Treatment",
       subtitle = "Effect of lapsing on vision impairment (log odds scale)",
       x = "DR Severity",
       y = "Log Odds Ratio",
       color = "Treatment Status") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "bottom") +
  scale_color_manual(values = c("No_Treatment" = "#2ca02c", 
                               "Any_Treatment" = "#d62728"))

# Create OR plot
p2 <- ggplot(plot_data, aes(x = DR_Severity, y = OR, 
                            fill = Treatment)) +
  geom_col(position = position_dodge(width = 0.8), alpha = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  labs(title = "Odds Ratios by Subgroup",
       subtitle = "Effect of lapsing on vision impairment",
       x = "DR Severity",
       y = "Odds Ratio",
       fill = "Treatment Status") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "bottom") +
  scale_fill_manual(values = c("No_Treatment" = "#2ca02c", 
                               "Any_Treatment" = "#d62728")) +
  geom_text(aes(label = sprintf("%.2f", OR)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3.5)

@@

>>>>

# Add OR confidence intervals (converting from log odds scale)
plot_data <- plot_data |>
  mutate(
    OR_CI_Lower = exp(CI_Lower),
    OR_CI_Upper = exp(CI_Upper)
  )

# Check for extreme values and cap if necessary
max_or <- max(c(plot_data$OR, plot_data$OR_CI_Upper), na.rm = TRUE)
if (max_or > 100) {
  warning(paste("Extreme OR values detected (max =", round(max_or, 1), "). Consider checking the model."))
  # Cap at reasonable values for visualization
  plot_data <- plot_data |>
    mutate(
      OR_display = pmin(OR, 20),
      OR_CI_Lower_display = pmax(OR_CI_Lower, 0.05),
      OR_CI_Upper_display = pmin(OR_CI_Upper, 20),
      capped = OR > 20 | OR_CI_Upper > 20
    )
} else {
  plot_data <- plot_data |>
    mutate(
      OR_display = OR,
      OR_CI_Lower_display = OR_CI_Lower,
      OR_CI_Upper_display = OR_CI_Upper,
      capped = FALSE
    )
}

# Create interaction plot with Odds Ratios (converted from log odds)
p1 <- ggplot(plot_data, aes(x = DR_Severity, y = OR_display, 
                            color = Treatment, group = Treatment)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  geom_errorbar(aes(ymin = OR_CI_Lower_display, ymax = OR_CI_Upper_display), 
                width = 0.1, size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  labs(title = "Three-Way Interaction: Lapse × DR Severity × Treatment",
       subtitle = ifelse(any(plot_data$capped), 
                        "Effect of lapsing (OR scale, capped at 20 for display)",
                        "Effect of lapsing on vision impairment (odds ratio scale)"),
       x = "DR Severity",
       y = "Odds Ratio",
       color = "Treatment Status") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "bottom") +
  scale_color_manual(values = c("No_Treatment" = "#2ca02c", 
                               "Any_Treatment" = "#d62728")) +
  coord_cartesian(ylim = c(0, max(plot_data$OR_CI_Upper_display) * 1.1))

# Create OR bar plot with error bars
p2 <- ggplot(plot_data, aes(x = DR_Severity, y = OR_display, 
                            fill = Treatment)) +
  geom_col(position = position_dodge(width = 0.8), alpha = 0.7) +
  geom_errorbar(aes(ymin = OR_CI_Lower_display, ymax = OR_CI_Upper_display),
                position = position_dodge(width = 0.8),
                width = 0.2, size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  labs(title = "Odds Ratios by Subgroup",
       subtitle = ifelse(any(plot_data$capped),
                        "Effect of lapsing (with 95% CI, capped at 20 for display)",
                        "Effect of lapsing on vision impairment (with 95% CI)"),
       x = "DR Severity",
       y = "Odds Ratio",
       fill = "Treatment Status") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "bottom") +
  scale_fill_manual(values = c("No_Treatment" = "#2ca02c", 
                               "Any_Treatment" = "#d62728")) +
  geom_text(aes(label = ifelse(capped, 
                               sprintf(">%.0f", OR_display),
                               sprintf("%.2f", OR))),
            position = position_dodge(width = 0.8),
            vjust = -1.5, size = 3.5) +
  coord_cartesian(ylim = c(0, max(plot_data$OR_CI_Upper_display) * 1.1))

@@

### Forest Plot of Average Marginal Effects

```{r ame-forest-plot}
#| fig-height: 6
#| fig-width: 10

# Prepare data for forest plot
forest_data <- ame_results |>
  mutate(
    # Create a combined label for the y-axis
    Subgroup = paste0(DR_Severity, " + ", Treatment),
    # Order for plotting (reverse for forest plot convention)
    Subgroup = factor(Subgroup, levels = rev(unique(Subgroup)))
  ) |>
  arrange(desc(Subgroup))

# Create forest plot
forest_plot <- ggplot(forest_data, aes(x = AME, y = Subgroup)) +
  # Add reference line at 0 (no effect)
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +

  # Add confidence intervals
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), 
                 height = 0.2, size = 0.8, color = "gray30") +

  # Add point estimates
  geom_point(aes(color = DR_Severity, shape = Treatment), size = 4) +

  # Labels and theme
  labs(
    title = "Average Marginal Effects of Lapsing on Vision Impairment",
    subtitle = "By DR severity and treatment status (with 95% CI)",
    x = "Average Marginal Effect (Probability Scale)",
    y = "",
    color = "DR Severity",
    shape = "Treatment"
  ) +

  # Theme adjustments
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.y = element_text(size = 11, hjust = 0),
    axis.text.x = element_text(size = 11),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.grid.major.y = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  ) +

  # Color scale
  scale_color_manual(values = c(
    "No_DR" = "#1f77b4",
    "NPDR" = "#ff7f0e", 
    "PDR" = "#d62728"
  )) +

  # Shape scale
  scale_shape_manual(values = c(
    "No_Treatment" = 16,  # Circle
    "Any_Treatment" = 17   # Triangle
  )) +

  # X-axis formatting
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1))

# Display the plot
print(forest_plot)

# Add text annotations with exact values
forest_plot_annotated <- forest_plot +
  geom_text(aes(label = sprintf("%.3f", AME)),
            nudge_y = 0.3, size = 3.5, color = "black")

print(forest_plot_annotated)
```
